---
title: "Baseballr"
author: "Matthew Adams"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, janitor, data.table, here, rlang, shinydashboard, shiny, DT, bslib, plotly, shinyWidgets, scales, tictoc)
```

```{r read data, warning=FALSE}
# Path to all files
data_path <- here("Data")

# List data files to compile
data_list <- list.files(path = data_path, pattern = "sportsref_download", recursive = TRUE, full.names = TRUE)

# Create data structure summary  of all files
tic()
type_summary <- map_dfr(data_list, function(file) {
  # Read only headers
  df <- read_csv(file, show_col_types = FALSE, n_max = 0)
  types <- map_chr(df, ~ class(.x)[1])
  
  tibble(
    file = file,
    n_vars = ncol(df)
  ) %>% 
    bind_cols(as_tibble_row(types))
})
toc()
```


```{r  }
# Create standard file format 
template_df <- read_csv(here("Data", "LSU", "sportsref_download_LSU_2025.csv")) %>% clean_names()
col_types_template <- map_chr(template_df, ~ class(.x)[1])

# Function to clean files
read_and_clean <- function(file, col_types) {
  df <- read_csv(file, show_col_types = FALSE) %>% 
    clean_names()
  
  for (col in names(col_types)) {  # Loop over every column name in the type template
  if (col %in% names(df)) {      # Only run if that column exists in the current file’s dataframe
    df[[col]] <- switch(         # Coerce the column to the desired type using the template
      col_types[[col]],          # This fetches the desired type for this column, e.g. "character"
      
      character = as.character(df[[col]]),  # If type is "character", convert it
      numeric   = as.numeric(df[[col]]),    # If type is "numeric", convert it
      integer   = as.integer(df[[col]]),    # If type is "integer", convert it
      logical   = as.logical(df[[col]]),    # If type is "logical", convert it
      
      df[[col]]  # default: if type doesn't match any above, leave it unchanged
    )
    }
  }
  
  # Extract school and year from filename
  file_base <- basename(file)
  school <- str_extract(file_base, "(?<=sportsref_download_).*?(?=_[0-9]{4})")
  year   <- str_extract(file_base, "\\d{4}")
  
  df <- df %>%
    mutate(
      school = school,
      year = as.integer(year)
    )
  
  return(df)
}

tic()
all_data <- suppressWarnings(
  map_dfr(data_list, ~ read_and_clean(.x, col_types_template))
)
toc()

# Final Data
df <- all_data %>% 
  mutate(name = str_replace_all(name, "[^A-Za-z ]", "")) %>% 
  filter(!is.na(rk)) %>% 
  filter(year > 2011) %>% 
  select(-notes) %>% 
  relocate(year, .before = everything()) %>% 
  relocate(school, .after = year)

unique(df$school)
```
Advanced statistics

1. Fielding Independent Pitching (FIP)
Estimates a pitcher’s performance independent of fielding.

2. K% (Strikeout Rate)
Shows how often a pitcher strikes out hitters, normalized by batters faced (BF).Unlike K/9, it doesn’t get skewed by low innings.

3. BB% (Walk Rate)

4. Strikeout-to-Walk Ratio

5. Batting Average on Balls in Play
It measures how often balls hit into the field of play turn into hits — excluding home runs, strikeouts, walks, and hit-by-pitches.


```{r computed stats}
# FIP variables
hr_weight <- 13
bb_weight <- 3
so_weight <- 2
fip_constant <- 3.1
fip_min_ip <- 10

# SO % variables
k_pct_min_bf <- 10

# BB % variables
bb_pct_min_bf <- 10

# BABIP variables
babip_min_ip <- 10

# Testing data
df <- df %>%
  mutate(
    fip = case_when(
      is.na(hr) | is.na(bb) | is.na(so) | is.na(ip) ~ NA_real_,
      ip < fip_min_ip ~ NA_real_,
      TRUE ~ round((((hr_weight * hr) + (bb_weight * bb) - (so_weight * so)) / ip + fip_constant), 3)
    ),
    
    k_pct = case_when(
      is.na(so) | is.na(bf) | bf < k_pct_min_bf ~ NA_real_,
      TRUE ~ round((so / bf), 3)
    ),
    
    bb_pct = case_when(
      is.na(bb) | is.na(bf) | bf < bb_pct_min_bf ~ NA_real_,
      TRUE ~ round((bb / bf), 3)
    ),
    
    babip_denominator = bf - so - bb - hbp - hr,
    
    babip = case_when(
      is.na(h) | is.na(hr) | is.na(so) | is.na(bb) | is.na(hbp) | is.na(bf) | is.na(ip) ~ NA_real_,
      ip < babip_min_ip ~ NA_real_,
      babip_denominator <= 0 ~ NA_real_,
      TRUE ~ round(((h - hr) / babip_denominator), 3)
    )
  )

```
This code calculates shrinkage-adjusted player ratings for ERA, SO/9 (strikeouts per 9 innings), and BB/9 (walks per 9 innings). The goal is to compare players fairly by reducing the impact of small sample sizes (i.e., low innings pitched).

We use a Bayesian-style formula to adjust each player's stat by blending their individual performance with the league average for that stat and year. This is controlled by a prior weight, which determines how heavily we weight the league average — the lower a player’s IP, the more their stat is “shrunk” toward that average.

shrunk_stat = (player_ip * player_stat + prior_weight * league_stat) / (player_ip + prior_weight)

Once shrunk values are calculated, each stat is rescaled to a 0–100 rating, where 100 represents the best performance for that stat in a given year.
```{r player rating}
# -------------------------------------------
# Define stats to rate and their direction
# -------------------------------------------

rating_stats <- tribble(
  ~stat,     ~higher_is_better, ~prior_weight,
  "era",     FALSE,             40,
  "fip",     FALSE,             30,
  "so9",     TRUE,              15,
  "bb9",     FALSE,             20,
  "hr9",     FALSE,             25,
  "whip",    FALSE,             25,
  "so_w",    TRUE,              10,
  "k_pct",   TRUE,              15,
  "bb_pct",  FALSE,             15
)

# -------------------------------------------
# Compute league averages per year
# -------------------------------------------

league_avgs <- df %>%
  filter(ip > 0) %>%
  group_by(year) %>%
  summarize(across(
    all_of(rating_stats$stat),
    ~ mean(.x, na.rm = TRUE),
    .names = "avg_{.col}"
  ), .groups = "drop")

# -------------------------------------------
# Join league averages back to player data
# -------------------------------------------

df_joined <- df %>%
  left_join(league_avgs, by = "year")

# -------------------------------------------
# Apply shrinkage for each stat
# -------------------------------------------

# Create named vector for prior weights
prior_weights <- rating_stats$prior_weight
names(prior_weights) <- rating_stats$stat

# Apply shrinkage using map and bind to df_joined
shrunk_stats <- map2_dfc(
  rating_stats$stat,
  prior_weights,
  ~ {
    stat <- .x
    prior <- .y
    avg_col <- paste0("avg_", stat)
    shrunk_col <- paste0("shrunk_", stat)
    
    tibble(!!shrunk_col := if_else(
      df_joined$ip > 0 &
        is.finite(df_joined[[stat]]) &
        is.finite(df_joined[[avg_col]]),
      (df_joined$ip * df_joined[[stat]] + prior * df_joined[[avg_col]]) / 
        (df_joined$ip + prior),
      NA_real_
    ))
  }
)

# Add to main dataframe
df_joined <- bind_cols(df_joined, shrunk_stats)


# -------------------------------------------
# Rescale to 0–100 ratings per year
# -------------------------------------------

# Add `rating_name` and `invert` flag to rating_stats
rating_stats <- rating_stats %>%
  mutate(
    shrunk_col = paste0("shrunk_", stat),
    rating_col = paste0(stat, "_rating"),
    invert = !higher_is_better
  )

# Function to compute scaled ratings for one group (i.e., one year)
rescale_ratings_by_year <- function(df_group) {
  # Map over stats to compute each rating column
  rating_cols <- pmap_dfc(
    rating_stats,
    function(stat, higher_is_better, prior_weight, shrunk_col, rating_col, invert) {
      values <- df_group[[shrunk_col]]
      values <- if (invert) -values else values
      
      tibble(!!rating_col := round(rescale(values, to = c(0, 100), na.rm = TRUE), 0))
    }
  )
  
  bind_cols(df_group, rating_cols)
}

# Apply per-year rescaling
df_ratings <- df_joined %>%
  group_by(year) %>%
  group_split() %>%
  map_dfr(rescale_ratings_by_year) %>%
  ungroup()


# -------------------------------------------
# Calculate composite rating (equal weights)
# -------------------------------------------

stat_weights <- c(
  era_rating    = 0.20,
  fip_rating    = 0.20,
  so9_rating    = 0.15,
  bb9_rating    = 0.10,
  hr9_rating    = 0.10,
  whip_rating   = 0.10,
  so_w_rating   = 0.05,
  k_pct_rating  = 0.05,
  bb_pct_rating = 0.05
)

# Only keep stats that exist in your df
valid_weights <- stat_weights[names(stat_weights) %in% colnames(df_ratings)]
rating_names <- names(valid_weights)

# Weighted composite
df_ratings <- df_ratings %>%
  mutate(
    composite_rating = round(
      rowSums(
        select(., all_of(rating_names)) * rep(valid_weights, each = nrow(.)),
        na.rm = TRUE
      ),
      0
    )
  )

# -------------------------------------------
# Final output
# -------------------------------------------

# Core identity columns
core_cols <- c("year", "school", "name", "ip", "composite_rating")

# Generate stat-wise column blocks in order
stat_column_blocks <- lapply(rating_stats$stat, function(stat) {
  c(
    stat,                           # raw stat
    paste0("avg_", stat),           # league avg
    paste0("shrunk_", stat),        # shrunk stat
    paste0(stat, "_rating")         # scaled 0–100 rating
  )
})

# Flatten list into character vector
stat_columns_ordered <- unlist(stat_column_blocks)

# Combine with identity columns
final_detailed_cols <- c(core_cols, stat_columns_ordered)

# Create final detailed output
test_all_ratings_detailed <- df_ratings %>%
  select(all_of(final_detailed_cols)) %>% 
  arrange(desc(composite_rating))

# Final user output columns
final_cols <- c("year", "school", "name", "ip", "composite_rating", rating_names)

# Final user output columns
test_all_ratings <- df_ratings %>%
  select(all_of(final_cols)) %>%
  arrange(desc(composite_rating))

```


```{r analysis 1}

df_analysis_1 <- df %>% 
  filter(ip > 0) %>% 
  group_by(year) %>% 
  summarise(across(
    where(is.numeric),
    list(
      mean = ~round(mean(.x, na.rm = TRUE), 3),
      median = ~round(median(.x, na.rm = TRUE), 3)),
    .names = "{.col}_{.fn}"
    ))

ggplot(df_analysis_1, aes(x = factor(year), y = so9_median, fill = factor(year))) +
  geom_col() +
  scale_fill_manual(values = rep(c("#582C83", "#FDB827"), length.out = length(unique(df_analysis_1$year)))) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "LSU Pitching Trends",
    subtitle = "Median SO/9 by Season",
    x = "Season",
    y = "Strikeouts per 9 Innings"
    )
```



