---
title: "Baseballr"
author: "Matthew Adams"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, janitor, data.table, here, rlang, shinydashboard, shiny, DT, bslib, plotly, shinyWidgets)
```


```{r data sourcing}
pitching_programs <- tribble(
  ~team,             ~notes,
  "Vanderbilt",        "Elite development; produced Price, Rocker, Leiter",
  "Florida",           "Consistent high-round draft picks; deep rotation",
  "UCLA",              "Pitching-first culture; Bauer, Cole, Garcia",
  "LSU",               "Strong pitching tradition in powerhouse program",
  "Arkansas",          "SEC standout; produced Kevin Kopps, Hagen Smith",
  "Mississippi State", "Pitching-focused with strong recent arms",
  "Texas A&M",         "Underrated pitching development",
  "Wake Forest",       "Analytical leader in biomechanics and pitching velocity",
  "Louisville",        "Solid college arms; Detmers, Funkhouser",
  "Stanford",          "West Coast development hub with pro-ready arms",
  "North Carolina",    "Historic pitching success; Bard, Harvey",
  "Oregon State",      "Efficient, well-coached staff; consistent results",
  "South Carolina",    "Known for elite pitching in championship era",
  "Tennessee",         "Recent pitching surge; Dollander, Beam",
  "TCU",               "Big 12 leader in pitching depth and development",
  "Florida State",     "Consistant NCAA Tournament appearences"
)

pitching_programs

```


```{r Baseball Reference, warning=FALSE}
# Path to all files
data_path <- here("Data")

# List data files to compile
data_list <- list.files(path = data_path, pattern = "sportsref_download", recursive = TRUE, full.names = TRUE)

# Function to read and clean a file
process_file <- function(file) {
  read_csv(file, show_col_types = FALSE) %>% 
    clean_names()
}

# Summary of each file's column types
type_summary <- map_dfr(data_list, function(file) {
  df <- process_file(file)
  types <- map_chr(df, ~ class(.x)[1])  # get class of each column
  
  tibble(
    file = file,
    n_vars = ncol(df)
  ) %>%
    bind_cols(as_tibble_row(types))
})

# Create standard file format 
template_df <- read_csv(here("Data", "LSU", "sportsref_download_LSU_2025.csv")) %>% clean_names()
col_types_template <- map_chr(template_df, ~ class(.x)[1])

read_and_clean <- function(file, col_types) {
  df <- read_csv(file, show_col_types = FALSE) %>% clean_names()
  
  for (col in names(col_types)) {  # Loop over every column name in the type template
  if (col %in% names(df)) {      # Only run if that column exists in the current file’s dataframe
    df[[col]] <- switch(         # Coerce the column to the desired type using the template
      col_types[[col]],          # This fetches the desired type for this column, e.g. "character"
      
      character = as.character(df[[col]]),  # If type is "character", convert it
      numeric   = as.numeric(df[[col]]),    # If type is "numeric", convert it
      integer   = as.integer(df[[col]]),    # If type is "integer", convert it
      logical   = as.logical(df[[col]]),    # If type is "logical", convert it
      
      df[[col]]  # default: if type doesn't match any above, leave it unchanged
    )
    }
  }
  
  # Extract school and year from filename
  file_base <- basename(file)
  
  school <- str_extract(file_base, "(?<=sportsref_download_)[A-Za-z]+")
  year   <- str_extract(file_base, "\\d{4}")
  
  df <- df %>%
    mutate(
      school = school,
      year = as.integer(year)
    )
  
  return(df)

}

all_data <- map_dfr(data_list, ~ read_and_clean(.x, col_types_template))

df <- all_data %>% 
  mutate(name = str_replace_all(name, "[^A-Za-z ]", "")) %>% 
  filter(!is.na(rk)) %>% 
  select(-notes) %>% 
  relocate(year, .before = everything()) %>% 
  relocate(school, .after = year)

```
Advanced statistics

1. Fielding Independent Pitching (FIP)
Estimates a pitcher’s performance independent of fielding.

2. K% (Strikeout Rate)
Shows how often a pitcher strikes out hitters, normalized by batters faced (BF).Unlike K/9, it doesn’t get skewed by low innings.

3. BB% (Walk Rate)

4. Strikeout-to-Walk Ratio

5. Batting Average on Balls in Play
It measures how often balls hit into the field of play turn into hits — excluding home runs, strikeouts, walks, and hit-by-pitches.


```{r computed stats}
# FIP variables
hr_weight <- 13
bb_weight <- 3
so_weight <- 2
fip_constant <- 3.1
fip_min_ip <- 10

# SO % variables
k_pct_min_bf <- 10

# BB % variables
bb_pct_min_bf <- 10

# K/BB % variables
k_bb_min_bb <- 1
k_bb_min_ip <- 10

# BABIP variables
babip_min_ip <- 10

# Testing data
test <- df %>%
  mutate(
    fip = case_when(
      is.na(hr) | is.na(bb) | is.na(so) | is.na(ip) ~ NA_real_,
      ip < fip_min_ip ~ NA_real_,
      TRUE ~ ((hr_weight * hr) + (bb_weight * bb) - (so_weight * so)) / ip + fip_constant
    ),
    
    k_pct = case_when(
      is.na(so) | is.na(bf) | bf < k_pct_min_bf ~ NA_real_,
      TRUE ~ so / bf
    ),
    
    bb_pct = case_when(
      is.na(bb) | is.na(bf) | bf < bb_pct_min_bf ~ NA_real_,
      TRUE ~ bb / bf
    ),
    
    k_bb = case_when(
      is.na(so) | is.na(bb) | is.na(ip) ~ NA_real_,
      ip < k_bb_min_ip ~ NA_real_,
      bb == 0 ~ NA_real_,  # or Inf, depending on your preference
      TRUE ~ so / bb
    ),
    
    babip_denominator = bf - so - bb - hbp - hr,
    
    babip = case_when(
      is.na(h) | is.na(hr) | is.na(so) | is.na(bb) | is.na(hbp) | is.na(bf) | is.na(ip) ~ NA_real_,
      ip < babip_min_ip ~ NA_real_,
      babip_denominator <= 0 ~ NA_real_,
      TRUE ~ (h - hr) / babip_denominator
    )
  )

```


```{r analysis 1}

df_analysis_1 <- df %>% 
  filter(ip > 0) %>% 
  group_by(year) %>% 
  summarise(across(
    where(is.numeric),
    list(
      mean = ~round(mean(.x, na.rm = TRUE), 3),
      median = ~round(median(.x, na.rm = TRUE), 3)),
    .names = "{.col}_{.fn}"
    ))

ggplot(df_analysis_1, aes(x = factor(year), y = so9_median, fill = factor(year))) +
  geom_col() +
  scale_fill_manual(values = rep(c("#582C83", "#FDB827"), length.out = length(unique(df_analysis_1$year)))) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "LSU Pitching Trends",
    subtitle = "Median SO/9 by Season",
    x = "Season",
    y = "Strikeouts per 9 Innings"
    )
```



